---
title: "R practice"
author: "NDP"
date: '2022-06-02'
output: html_document
---

## Intro to markdown

Markdown allows you to do things like **bold**, _italic_, 'codefont' and more!

make a list with astrisks like 
* bold with double astrisks

* italics with underscores

* code- type with backticks

* you can aslo use hyphens instead of astrisks

For a number list:

1. 1\. bold....

2. 1\. italic...

3. 1\. code...

Hyperlink are like [hyperlinkedtext](https://mylink.org)

```{r load_data}
gap <-gapminder
```

```{r make-plot}
plot(lifeExp ~ year , data=gap)
temp <- C(1,2,3)
```

```{r make_cats}
cats <-data.frame(coat = c("calico", "black", "tabby"), weight= c(2.1, 5.0, 3.2), likes_string = c(1, 0, 1))
write.csv(x = cats, file = "data/feline-data.csv", row.names = FALSE)
```

```{r read_cats}
cats2 <- read.csv("data/feline-data.csv",stringsAsFactors = TRUE)

#creating vectors and coercing type

```{r}
my_vector <- vector(made="chractor", length = 3)
combine_vector <- C(1,2,"bananas")
```
change names

```{r}
names(cats)[3] <- "likes_hunting"
```

rows
```{r}
age <- c(2,3,5)
cats <- cbind(cats,age)
newrow <- list("tortoiseshell", 3,3,TRUE, 9)
```

```{r}
human_age <- cats$age * 7
human_age <- factor(human_age)
human_age <- as.numeric(as.character(human_age))
human_age / 7
```

```{r}
cats$coat <- as.factor(cats$coat)
cats <- rbind(cats,c("white", 3,4,0,1))
#remove row
cats <- cats[-5,]
```


```{r}
library(readxl)
#download at https://github.com/ndporter/dspg-r-2022/raw/gh-pages/data/gapminder_data.xlsx
gapminder <- read_excel("data/gapminder_data.xlsx")
```

```{r}
summary(gapminder)
```

```{r}
length(gapminder)
colnames(gapminder)
head(gapminder, 6)
tail(gapminder)
gapminder[sample(nrow(gapminder),10),]

```
```{r}
x <- c(5,4,6,2,7,1,4,8,7,5)
x[1:3]
x[c(1,1,3)]
x-c(1,5)
x[1:3]
x <- x[-4]
x <- c(5,4,6,2,7,1,4,8,7,5)
x<7

```
```{r}
seAsia <- c("Myanmar", "Thailand", "Cambodia", "Vietnam", "Loas")
gapminder[which(gapminder$country %in% seAsia),]

```
```{r}
gap_57 <- gapminder[gapminder$year==1959 | gapminder$year==1970, ]
```


```{r}
library(ggplot2)
ggplot(data =gapminder, mapping =aes(x = year, y = lifeExp)) + geom_point()
```
```{r}
library(ggplot2)
ggplot(data =gapminder, mapping =aes(x = year, y = lifeExp, color=continent)) +geom_line() 
```
```{r}
library(ggplot2)
ggplot(data =gapminder, mapping =aes(x = year, y = lifeExp, by=country, )) +geom_line(mapping = aes(color=continent)) + geom_point(aes(color="blue"))
```
```{r}
ggplot(gapminder, mapping = aes(x=gdpPercap, y=lifeExp)) + geom_point() +scale_x_log10() 
```
```{r}
logged_points <- ggplot(gapminder, mapping = aes(x=gdpPercap, y=lifeExp)) + geom_point() +scale_x_log10() 


logged_points +geom_smooth(method= loess)
```
```{r}
americas <- gapminder[gapminder$continent=="Americas",]
ggplot(data=americas, mapping =aes(x=year, y=lifeExp)) + geom_line()+ facet_wrap( ~ country) +
theme(axis.text.x = element_text(angle = 45))
```
```{r}
write.csv(file= "clean_data/americas_gap.csv", x=americas, row.names = F)
```

```{r}
#mean(gapfinder[gapminder$continent=="Africa"], "gdpPercap")
```

```{r}
library("dplyr")
year_contry_gdp <- select(gapminder, year, country, gdpPercap)
str(year_contry_gdp)
smaller_gdp <- gapminder %>% select(-continent) %>% rename(gdp_per_capita=gdpPercap)
```



